Efficient UQ for chemical systems using active \author{Manav Vohra, Alen Alexanderian, Hayley Guy, Sankaran Mahadevan} subspaces

### 2018-08-13 Paper/intro.tex
..

Time evolution of a chemically reacting system is largely dependent upon rate
constants associated with individual reactions. The rate constants are
typically assumed to exhibit a certain correlation with temperature (e.g.
Arrhenius-type). Hence, accurate specification of the rate-controlling
parameters is critical to the fidelity of simulations. However, in practical
applications, the parameters are either specified using expert knowledge or
estimated based on a regression-fit to a set of sparse and noisy
data~\cite{Burnham:1987, Burnham:1988, Vohra:2011, Sarathy:2012}.
Consequently, the underlying uncertainty associated with the parameters and
therefore the simulation is typically ignored. Nevertheless, the inclusion of
uncertainty due to corrupted calibration data, parametric uncertainty,
and the discrepancy between predictions and experiments
due to model inadequacy has gained prominence in recent
years~\cite{Vohra:2014, Vohra:2017, Morrison:2018, Hantouche:2018}. 
However, in complex mechanisms involving a large number of reactions,
characterizing the propagation of uncertainty from a large set of inputs to
the model output becomes remarkably challenging due to associated
computational effort. 
A major focus of this paper pertains to the implementation of a robust
framework that aims to identify `important' directions in the input space
that predominantly captures the variability in the model output. These
directions, referred to as the \textit{active subspaces}~\cite{Constantine:2015},
constitute the dominant
eigenspace of a matrix derived from the gradient information of the model
output with respect to an input. The active subspace methodology thus
focuses on reducing the dimensionality of the problem.

The application problem considered in this work is the
H$_2$/O$_2$ reaction mechanism~\cite{Yetter:1991}. This mechanism has gained
a lot of attention as a potential source of clean energy for
locomotive applications~\cite{Das:1996}, and more recently in fuel cells~\cite{Loges:2008}. 
The mechanism involves 19
reactions including chain reactions, dissociation/recombination reactions, and
formation and consumption of intermediate species. For each reaction in
table~\ref{tab:kinetics}, the reaction rate is assumed to show an Arrhenius
correlation with temperature:
%
\be
k_i(T) = A_iT^{n_i}\exp(-E_{a,i}/RT), 
\label{eq:rate}
\ee
%
where $A_i$ is the pre-exponent, $n_i$ is the index of $T$, $E_{a,i}$ is the
activation energy corresponding to the $i^{th}$ reaction, and $R$ is the
universal gas constant.

@@

We are interested in investigating the impact of uncertainty in the
rate-controlling parameters, pre-exponent~($A_i$) and the activation
energy~($E_{a,i}$) on the ignition delay. The $A_i$'s are considered to be
uniformly distributed about their nominal estimates provided
in~\cite{Yetter:1991}.  The total number of uncertain inputs is 33 which makes
the present problem high-dimensional and computationally challenging. 
To address this challenge, we focus our efforts on reducing the dimensionality
of the problem by computing the active subspace using
two computational strategies: a gradient-based approach and  
a gradient-free approach. These strategies are based on algorithms 1.1 and 1.2 
in~\cite{Constantine:2015}. As mentioned earlier, the gradient information of the model
output is needed in both cases. However, in the case of gradient-free approach,
the gradient is approximated using a linear regression-fit to the model output.
Hence, the amount of computational effort is reduced in this case. Both strategies
are implemented in an iterative manner i.e. a new set of model evaluations are
generated, and gradient evaluations are performed at each iteration until 
a reasonably accurate approximation to the active subspace is achieved, as
discussed later in Section~\ref{sec:method}.

An alternate approach to dimension reduction involves computing the global sensitivity
measures associated with uncertain inputs of a model. Depending upon the estimates of
the sensitivity measures, only the important inputs are varied for the purpose of
uncertainty quantification (UQ). The
Sobol' index is a commonly used global sensitivity measure~\cite{Sobol:2001}. However, obtaining
converged estimates of the Sobol' indices typically requires tens of thousands of model
runs and is therefore impractical in situations involving intensive model simulations.
Recently, the derivative-based global sensitivity measures~(DGSMs) were 
developed to approximate the upper bound on the Sobol' indices with much fewer
computations~\cite{Sobol:2009, Lamboni:2013}. In Section~\ref{sec:assa}, we show
that the DGSMs can in fact be approximated using the components of the eigenvectors
in the active subspace by estimating the so-called activity 
scores~\cite{Diaz:2016,Constantine:2017}.

Main contributions of this paper are as follows: (1) Mathematical linkages between the
global sensitivity measures: Total Sobol' index, DGSMs, and the activity scores are
established; (2) An efficient gradient-free iterative approach using regression-based local linear
approximation of the model output is shown to yield consistent results with the
gradient-based approach. The two approaches were implemented to compute
the active subspace for a 19-dimensional H$_2$/O$_2$ kinetics problem with 
uncertain pre-exponents in the rate law for individual reactions. (3) The gradient-free
approach is finally implemented to discover an active subspace for the 33-dimensional
H$_2$/O$_2$ reaction kinetics application in which case the pre-exponents as well
as the activation energies are considered to be uncertain.

 The remaining of this paper is organized as follows. In section~\ref{sub:ac}, a brief theoretical background on
 the active subspace methodology is provided. In section~\ref{sub:gsa}, it is shown that the activity
 scores provide a reasonable approximation to the DGSMs especially in a high-dimensional setting. 
 Additionally, a relationship between the three global sensitivity measures, namely, the activity scores,
 DGSMs, and the total Sobol' indices is established. In section~\ref{sec:method}, a systematic framework
 for computing the active subspace using the gradient-based, and the gradient-free approaches is presented. 
The two approaches are implemented to the H$_2$/O$_2$ reaction kinetics problem whereby only
the $A_i$'s are considered as uncertain. Our findings illustrate the consistency in the results based on
the two approaches thereby encouraging the use of the gradient-free approach in high-dimensional settings.
 In section~\ref{sec:app},
the gradient-free approach is implemented to the 33-dimensional
H$_2$/O$_2$ reaction kinetics application by accounting for the uncertainty in $E_{a,i}$'s as well. The
resulting active subspace is found to be 1-dimensional indicating enormous potential for reduction in
computational effort pertaining to UQ, in this case. 
Finally, we summarize the results and include a discussion on this work in section~\ref{sec:conc}.

>>>>

Time evolution of a chemically reacting system is largely dependent upon rate
constants associated with individual reactions. The rate constants are
typically assumed to exhibit a certain correlation with temperature (e.g.
Arrhenius-type). Hence, accurate specification of the rate-controlling
parameters is critical to the fidelity of simulations. However, in practical
applications, the parameters are either specified using expert knowledge or
estimated based on a regression-fit to a set of sparse and noisy
data~\cite{Burnham:1987, Burnham:1988, Vohra:2011, Sarathy:2012}.
Consequently, the underlying uncertainty associated with the parameters and
therefore the simulation is typically ignored. Nevertheless, the inclusion of
uncertainty due to corrupted calibration data, parametric uncertainty,
and the discrepancy between predictions and experiments
due to model inadequacy has gained prominence in recent
years~\cite{Vohra:2014, Vohra:2017, Morrison:2018, Hantouche:2018}. 
However, in complex mechanisms involving a large number of reactions,
characterizing the propagation of uncertainty from a large set of inputs to
the model output becomes remarkably challenging due to associated
computational effort. 
A major focus of this paper pertains to the implementation of a robust
framework that aims to identify `important' directions in the input space
that predominantly captures the variability in the model output. These
directions, referred to as the \textit{active subspaces}~\cite{Constantine:2015},
constitute the dominant
eigenspace of a matrix derived from the gradient information of the model
output with respect to an input. The active subspace methodology thus
focuses on reducing the dimensionality of the problem, and hence the 
computational effort associated with uncertainty propagation.

The application problem considered in this work is the
H$_2$/O$_2$ reaction mechanism~\cite{Yetter:1991}. This mechanism has gained
a lot of attention as a potential source of clean energy for
locomotive applications~\cite{Das:1996}, and more recently in fuel
 cells~\cite{Loges:2008,Cosnier:2016}. 
The mechanism involves 19
reactions including chain reactions, dissociation/recombination reactions, and
formation and consumption of intermediate species. For each reaction in
table~\ref{tab:kinetics}, the reaction rate is assumed to show an Arrhenius
correlation with temperature:
%
\be
k_i(T) = A_iT^{n_i}\exp(-E_{a,i}/RT), 
\label{eq:rate}
\ee
%
where $A_i$ is the pre-exponent, $n_i$ is the index of $T$, $E_{a,i}$ is the
activation energy corresponding to the $i^{th}$ reaction, and $R$ is the
universal gas constant.

@@

We are interested in investigating the impact of uncertainty in the
rate-controlling parameters, pre-exponent~($A_i$) and the activation
energy~($E_{a,i}$) on the ignition delay. The $A_i$'s associated with all
reactions and the $E_{a,i}$'s with non-zero nominal estimates
are considered to be uniformly distributed about their nominal estimates provided
in~\cite{Yetter:1991}.  The total number of uncertain inputs is 33 which makes
the present problem high-dimensional and computationally challenging. 
To address this challenge, we focus our efforts on reducing the dimensionality
of the problem by computing the aforementioned active subspace using
an approach that involves computation of gradients of a model output with
respect to individual parameters. Estimating the
gradients can however be computationally challenging in high-dimensional settings. 
Hence, in this work, we further explore the suitability of a gradient-free
approach that involves a regression-based local linear approximation of
the model output. It must be noted that since the gradient-free approach
relies on approximating the model gradients, it is expected to be less
robust than the gradient-based approach. The two approaches are based on
algorithms 1.1 and 1.2 in~\cite{Constantine:2015}. Both strategies
are implemented in an iterative manner i.e. a new set of model evaluations are
generated, and gradient evaluations are performed at each iteration until 
a reasonably accurate approximation to the active subspace is achieved, as
discussed later in Section~\ref{sec:method}.

An alternate approach to dimension reduction involves computing the global sensitivity
measures associated with uncertain inputs of a model. Depending upon the estimates of
the sensitivity measures, only the important inputs are varied for the purpose of
uncertainty quantification (UQ). The
Sobol' index is a commonly used global sensitivity measure~\cite{Sobol:2001}. However, obtaining
converged estimates of the Sobol' indices typically requires tens of thousands of model
runs and is therefore impractical in situations involving intensive model simulations and/or a
large number of inputs.
Recently, the derivative-based global sensitivity measures~(DGSMs) were 
developed to approximate the upper bound on the Sobol' indices with much fewer
computations~\cite{Sobol:2009, Lamboni:2013}. In Section~\ref{sec:assa}, we show
that the DGSMs can in fact be approximated using the components of the eigenvectors
in the active subspace by estimating the so-called activity 
scores~\cite{Diaz:2016,Constantine:2017}.

Main contributions of this paper are as follows: (1) Mathematical linkages between the
global sensitivity measures: Total Sobol' index, DGSMs, and the activity scores are
established; (2) An initial comparison of the active subspace and the global sensitivity
measures, computed using the gradient-based and the gradient-free approaches in
a 19-dimensional H$_2$/O$_2$ kinetics problem is provided. In this case, only the $A_i$'s are
considered to be uncertain. (3) Active subspace discovery in the high-dimensional
H$_2$/O$_2$ kinetics problem involving 33 uncertain inputs.

 The remaining of this paper is organized as follows. In section~\ref{sub:ac}, a brief theoretical background on
 the active subspace methodology is provided. In section~\ref{sub:gsa}, it is shown that the activity
 scores provide a reasonable approximation to the DGSMs especially in a high-dimensional setting. 
 Additionally, a relationship between the three global sensitivity measures, namely, the activity scores,
 DGSMs, and the total Sobol' indices is established. In section~\ref{sec:method}, a systematic framework
 for computing the active subspace using an iterative gradient-based approach is provided. Numerical results
 based on the gradient-free approach are compared with those obtained using the gradient-based approach.
 The active subspace is computed for the H$_2$/O$_2$ reaction kinetics problem whereby only
the $A_i$'s are considered as uncertain. In this case, the two approaches are observed to yield consistent
results. In section~\ref{sec:app}, the gradient-based approach is further implemented to compute the
active subspace for the high-dimensional H$_2$/O$_2$ reaction kinetics problem. The resulting active
subspace is found to be 1-dimensional thereby indicating enormous potential for computational savings
in this application. The active subspace is also computed using the gradient-free approach. It is observed
that although the uncertainty in the ignition delay is reasonably approximated, the gradient-free approach
fails to capture the global sensitivity of all the uncertain inputs. Hence, a trade-off between efficiency and
accuracy between the two approaches is observed in this case. 
Finally, a summary and discussion based on our findings is included in section~\ref{sec:conc}.
